---
title: "CHOK1 DP12 cell line scRNASeq"
author: Clarke Lab. NIBRT
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
  html_notebook: 
    df_print: paged
    smooth_scroll: yes
    theme: cerulean
    toc: yes
    toc_float: yes
    highlight: haddock
    number_sections: yes
    fig_caption: yes
    fig_width: 5
    fig_height: 5
  pdf_document:
    toc: yes
---
```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(width = 75)
```
***
# Introduction

This R notebook enables the reproduction of the analysis described in:  


**Tracing production instability in a clonally-derived CHO cell line using single cell transcriptomics**


In this study, the four biological replicates of a clonally derived CHO cell line were
sampled at 72hrs post-seeding and analysed on the BD Rhapsody system whole transcriptome analysis.  The FASTQ
data were converted to a cell gene count matrix using the [Seven Bridges Genomics](https://www.sevenbridges.com/) BD
Rhapsody WTA pipeline and combined into a single sample. 

## Raw Data Availibility

SRA ID: PRJNA661407

# Preparation 

## Load packages
```{r load_packages, echo=TRUE, message=FALSE, warning=FALSE}
library(monocle)
library(plyr)
library(dplyr)
library(ggExtra)
library(reshape)
library(colorRamps)
library(stringr)
library(Seurat)
library(BiocParallel)
library(ggpubr)
library(DT)
library(WebGestaltR)
library(ggalt)
library(biomaRt)
library(patchwork)
library(viridis)
library(gridExtra)
library(grid)
library(gganimate)
library(hrbrthemes)
library(jcolors)
library(cowplot)
library(data.table)
library(GGally)
library(ggplotify)
library(stringr)
library(writexl)
```

## Set seed
```{r seed_for_reproducibility}
# set a seed for reproducibility
set.seed(38)
```

## Plotting utilites 
```{r plot_setup}
# ggplot theme
theme_ms <- function(base_size = 10, base_family = "Helvetica") {
  library(grid)
  (theme_bw(base_size = base_size, base_family = base_family) +
    theme(
      text = element_text(color = "black"),
      axis.title = element_text(face = "bold", size = rel(1)),
      plot.title = element_text(face = "bold", size = rel(1)),
      axis.text = element_text(size = rel(1), color = "black"),
      legend.title = element_text(face = "bold"),
      legend.text = element_text(face = "bold"),
      legend.background = element_rect(fill = "transparent"),
      legend.key.size = unit(0.8, "lines"),
      panel.border = element_rect(color = "black", size = 1),
      panel.grid = element_blank()
    ))
}

# pallette for groups
colour.set <- c(
  "#29BF12", "#009ACD", "#B1740F", "#DB2763", "#9A32CD",
  "#628395", "#17377A", "#E5C616", "#FF7844", "#E48F1B"
)

# color for expression values
color.gradient <- scale_color_gradientn(colors = matlab.like(200), 
                                        trans = "log10")

# create a directory for saving figures
figsdir <- "./figures/"
if (!dir.exists(figsdir)) {
  dir.create(figsdir)
}

# function to save figures
saveManuscriptPlot <- function(p, width, height) {
  figfile <- file.path(figsdir, sprintf(
    "%s.png",
    gsub("\\.", "_", deparse(substitute(p)))
  ))
  ggsave(figfile, plot = p, width = width, height = height, units = "in", device = "png")
}
```

## Calculate relative expression
Create a function to determine the relative expression of a gene from a Monocle cell dataset. When the expression of a gene is than the detection limit, the gene is equal to the detection limit.
```{r}
monocle_relative_expression <- function(cds, gene) {
  normalized_expression <- t(exprs(cds[gene, ])) / sizeFactors(cds)
  normalized_expression[normalized_expression < cds@lowerDetectionLimit] <- cds@lowerDetectionLimit
  return(normalized_expression)
}
```

# Import raw data
## scRNASeq 
We import data from the [Seven Bridges Genomics](https://www.sevenbridges.com/) (SBG) platform for analysis in R. The UMI data for each replicate is stored in a separate CSV file with each row representing an individual cell, the column represents a gene. In this experiment, 4 biological replicates for the CHOK1 DP12 cell line were analysed using Rhapsody WTA. We create a batch ID than be used to regress any technical variation due to batch. We determine the genes detected **in at least one replicate**. If a gene is not detected in a replicate we assign a 0 UMI value for that replicate. The replicates are then combined in a single matrix.

### SBG output
```{r load_data}
raw_umi_r1 <- read.table("scrnaseq_data/CHOK1DP12_r1.csv", sep = ",", header = T, row.names = 1, check.names = FALSE)
raw_umi_r1_batch <- rep("rep1", dim(raw_umi_r1)[1])

raw_umi_r2 <- read.table("scrnaseq_data/CHOK1DP12_r2.csv", sep = ",", header = T, row.names = 1, check.names = FALSE)
raw_umi_r2_batch <- rep("rep2", dim(raw_umi_r2)[1])

raw_umi_r3 <- read.table("scrnaseq_data/CHOK1DP12_r3.csv", sep = ",", header = T, row.names = 1, check.names = FALSE)
raw_umi_r3_batch <- rep("rep3", dim(raw_umi_r3)[1])

raw_umi_r4 <- read.table("scrnaseq_data/CHOK1DP12_r4.csv", sep = ",", header = T, row.names = 1, check.names = FALSE)
raw_umi_r4_batch <- rep("rep4", dim(raw_umi_r4)[1])

# determine the genes detected in at least 1 replicate
detected_genes <- unique(c(colnames(raw_umi_r1), colnames(raw_umi_r2), colnames(raw_umi_r3), colnames(raw_umi_r4)))
unique_genes <- unique(detected_genes)

# add a 0 column if gene is not detected in that replicate
raw_umi_r1[, setdiff(detected_genes, colnames(raw_umi_r1))] <- 0

raw_umi_r2[, setdiff(detected_genes, colnames(raw_umi_r2))] <- 0

raw_umi_r3[, setdiff(detected_genes, colnames(raw_umi_r3))] <- 0

raw_umi_r4[, setdiff(detected_genes, colnames(raw_umi_r4))] <- 0

raw_umi <- rbind(raw_umi_r1, raw_umi_r2, raw_umi_r3, raw_umi_r4)

# add text to cell label for Monocle compatability later
rownames(raw_umi) <- str_c("cell_", rownames(raw_umi)) # add text to ensure comparability with monocle plotting later

raw_umi_batch <- c(raw_umi_r1_batch, raw_umi_r2_batch, raw_umi_r3_batch, raw_umi_r4_batch) # record which Rhapsody cartridge the data came from
```
### Create Monocle object
The primary analysis package for this analysis Monocle v2. To use this package we must first create a *cell dataset* (CDS) object containing the raw Rhapsody data. During this process we assign a detection limit along with the expressionFamily - in this case *negbinomial* as appropriate for UMI data.  
```{r create_CDS}
pheno.data <- rep("cho", dim(raw_umi)[1])
pheno.data.df <- data.frame(type = pheno.data, batch = raw_umi_batch)
rownames(pheno.data.df) <- rownames(raw_umi)
pd <- new("AnnotatedDataFrame", data = pheno.data.df)

f.data <- colnames(raw_umi)
f.data.df <- data.frame(type = f.data) # Must be data frame object
rownames(f.data.df) <- colnames(raw_umi)
fd <- new("AnnotatedDataFrame",
  data = f.data.df
)

cells <- new("CellDataSet",
  exprs = t(raw_umi),
  phenoData = pd,
  featureData = fd,
  lowerDetectionLimit = 0.5,
  expressionFamily = negbinomial()
)

paste("Raw data for", dim(cells)[2], "and", dim(cells)[1], " genes captured")
```

## bulk RNASeq
To compare a matched bulkRNASeq dataset to our single cell data were calculated a gene-level transcripts per million (TPM) value using Kallisto.
```{r}
# import the TPM results from Kalisto for the four matched bulk samples
bulk_tpm <- read.table("bulkrnaseq_data/gene_mab_tpms_all_samples.tsv", header = T, row.names = 1)

# separate the bulk samples from the Kalisto TPM output and convert to log(tpm+1)
bulk_tpm_r1 <- log((bulk_tpm[, 2]) + 1)
names(bulk_tpm_r1) <- rownames(bulk_tpm)

bulk_tpm_r2 <- log((bulk_tpm[, 3]) + 1)
names(bulk_tpm_r2) <- rownames(bulk_tpm)

bulk_tpm_r3 <- log((bulk_tpm[, 4]) + 1)
names(bulk_tpm_r3) <- rownames(bulk_tpm)

bulk_tpm_r4 <- log((bulk_tpm[, 5]) + 1)
names(bulk_tpm_r4) <- rownames(bulk_tpm)

# determine the correlation between each of the scRNASeq samples
bulk_tpm <- data.frame(cbind(
  bulk_tpm_r1,
  bulk_tpm_r2,
  bulk_tpm_r3,
  bulk_tpm_r4
))

# rename the columns for each replicate
colnames(bulk_tpm) <- c("Replicate 1", "Replicate 2", "Replicate 3", "Replicate 4")
```


# Replicate comparison
## scRNASeq
To compare to the bulk RNASeq data we merge the scRNASeq to  pseudoBulk profiles, 1 for each replicate and convert expression to TPM. We then determine the agreement between the replicates
```{r bulk_comparison, fig.height=6, fig.width=6, message=FALSE}
# separate the scRNA by Rhapsody cartridge from the monocle CDS
# sum the UMI counts and convert to log(TPM+1)

pseudocounts_r1 <- rowSums(exprs(cells)[, pData(cells)$batch == "rep1"])
pseudocounts_r2 <- rowSums(exprs(cells)[, pData(cells)$batch == "rep2"])
pseudocounts_r3 <- rowSums(exprs(cells)[, pData(cells)$batch == "rep3"])
pseudocounts_r4 <- rowSums(exprs(cells)[, pData(cells)$batch == "rep4"])

# create logged pseduobulk TPM
norm_pseudocounts_r1 <- log(((pseudocounts_r1) / sum(pseudocounts_r1 / 1e6)) + 1)
norm_pseudocounts_r2 <- log(((pseudocounts_r2) / sum(pseudocounts_r2 / 1e6)) + 1)
norm_pseudocounts_r3 <- log(((pseudocounts_r3) / sum(pseudocounts_r3 / 1e6)) + 1)
norm_pseudocounts_r4 <- log(((pseudocounts_r4) / sum(pseudocounts_r4 / 1e6)) + 1)

# determine the correlation between each of the scRNASeq samples
norm_pseudocounts <- data.frame(cbind(
  norm_pseudocounts_r1,
  norm_pseudocounts_r2,
  norm_pseudocounts_r3,
  norm_pseudocounts_r4
))

colnames(norm_pseudocounts) <- c("Replicate 1", "Replicate 2", "Replicate 3", "Replicate 4")

fig.cor.scRNASeq <- ggpairs(norm_pseudocounts,
  lower = list(
    continuous = wrap("points",
      alpha = 0.3,
      size = 0.01
    ),
    combo = wrap("dot",
      alpha = 0.4,
      size = 0.2
    )
  )
) +
  theme_ms() +
  theme(panel.spacing = grid::unit(1, "lines"))

# save the plot for Supplementary Figure 1
s.fig1 <- fig.cor.scRNASeq
saveManuscriptPlot(s.fig1, 8, 8)
s.fig1
```

## bulk RNASeq
Determine the agreement between the bulkRNASeq replicates 
```{r}
fig.cor.bulkRNASeq <- ggpairs(bulk_tpm,
  lower = list(
    continuous = wrap("points",
      alpha = 0.3,
      size = 0.01
    ),
    combo = wrap("dot", alpha = 0.4, size = 0.2)
  )
) +
  theme_ms() +
  theme(panel.spacing = grid::unit(1, "lines"))

s.fig2 <- fig.cor.bulkRNASeq
saveManuscriptPlot(s.fig2, 8, 8)
s.fig2
```
# scRNASeq v bulkRNASeq
Comparison of each scRNASeq sample to it’s respective matched bulkRNASeq sample.

## Map gene IDs
```{r}
# annotate the genes with biomaRt
ensembl <- useMart(
  host = "http://sep2019.archive.ensembl.org",
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "cgpicr_gene_ensembl"
)

picr_annotation <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "entrezgene_id",
    "external_gene_name",
    "description",
    "gene_biotype"
  ),
  filters = c("ensembl_gene_id"),
  values = rownames(bulk_tpm),
  mart = ensembl,
  uniqueRows = TRUE
)

# the seven bridges genomics pipeline outputs both symbols and ens gene ids
# get the ids that are matched to gene symbols in the cds
symbols_cds <- as.character(fData(cells)$type[grep("ENSCGR", fData(cells)$type, invert = TRUE)])
matched_to_cds_symbols <- picr_annotation[as.character(picr_annotation$external_gene_name) %in% fData(cells)$type, ]
matched_to_cds_symbols <- unique(setDT(matched_to_cds_symbols), by = c("ensembl_gene_id", "external_gene_name"))
matched_to_cds_symbols <- matched_to_cds_symbols[, c(1, 3)]

# ensembl ids are common to bulk and scRNASeq
ensembl_cds <- as.character(fData(cells)$type[grep("ENSCGR", fData(cells)$type)])

# the PICR geneome has no MT chromosome, add the ensembl ids for the CHOK1 mtDNA genome
mito.genes <- c(
  "ND1", "ND2", "COX1", "COX2", "ATP8", "ATP6",
  "COX3", "ND3", "ND4L", "ND4", "ND5", "ND6", "CYTB"
)

mito.ensembl <- c(
  "ENSCGRG00000000006", "ENSCGRG00000000010", "ENSCGRG00000000016", "ENSCGRG00000000019",
  "ENSCGRG00000000021", "ENSCGRG00000000022", "ENSCGRG00000000023", "ENSCGRG00000000025",
  "ENSCGRG00000000027", "ENSCGRG00000000028", "ENSCGRG00000000032", "ENSCGRG00000000033",
  "ENSCGRG00000000035"
)

# create the ids for both datasets and include the mAb genes
cds_ids <- c(matched_to_cds_symbols$external_gene_name, ensembl_cds, "heavy_chain", "light_chain", mito.genes)
bulk_ids <- c(matched_to_cds_symbols$ensembl_gene_id, ensembl_cds, "heavy_chain", "light_chain", mito.ensembl)
```

## Correlation between single cell and bulk RNASeq data
```{r}
# determine the density of points to colour plot
density_r1 <- densCols(bulk_tpm_r1[bulk_ids], norm_pseudocounts_r1[cds_ids],
  colramp = colorRampPalette(rev(rainbow(10, end = 4 / 6))), nbin = 64
)
density_r2 <- densCols(bulk_tpm_r2[bulk_ids], norm_pseudocounts_r2[cds_ids],
  colramp = colorRampPalette(rev(rainbow(10, end = 4 / 6))), nbin = 64
)
density_r3 <- densCols(bulk_tpm_r3[bulk_ids], norm_pseudocounts_r3[cds_ids],
  colramp = colorRampPalette(rev(rainbow(10, end = 4 / 6))), nbin = 64
)
density_r4 <- densCols(bulk_tpm_r4[bulk_ids], norm_pseudocounts_r4[cds_ids],
  colramp = colorRampPalette(rev(rainbow(10, end = 4 / 6))), nbin = 64
)

# create the data frame for the plot
bulk_sc_cor_df <- data.frame(
  replicate = c(
    rep("Replicate 1", length(bulk_ids)),
    rep("Replicate 2", length(bulk_ids)),
    rep("Replicate 3", length(bulk_ids)),
    rep("Replicate 4", length(bulk_ids))
  ),
  bulk_tpm = c(
    bulk_tpm_r1[bulk_ids],
    bulk_tpm_r2[bulk_ids],
    bulk_tpm_r3[bulk_ids],
    bulk_tpm_r4[bulk_ids]
  ),
  scrnaseq_tpm = c(
    norm_pseudocounts_r1[cds_ids],
    norm_pseudocounts_r1[cds_ids],
    norm_pseudocounts_r1[cds_ids],
    norm_pseudocounts_r1[cds_ids]
  ),
  density = c(
    density_r1,
    density_r2,
    density_r3,
    density_r4
  )
)

fig.scatter.sc.v.bulk <- bulk_sc_cor_df %>%
  ggplot() +
  geom_point(aes(x = bulk_tpm, y = scrnaseq_tpm, col = density), size = 0.1) +
  scale_color_identity() +
  stat_cor(method = "pearson", aes(x = bulk_tpm, y = scrnaseq_tpm), inherit.aes = TRUE) +
  facet_wrap(~replicate) +
  theme_ms() +
  labs(x = "Bulk RNASeq Log(TPM+1)", y = "Pseudobulk RNASeq Log(TPM+1)") +
  theme(strip.background = element_blank())

s.fig3 <- fig.scatter.sc.v.bulk
saveManuscriptPlot(s.fig3, 8, 8)
s.fig3
```

# scRNASeq data preprocessing
The first stage of preprocessing in the Monocle workflow aims to eliminate poor quality cells from further analysis. In this section we calculate metrics to enable filtering of poor quality cells from the analysis - total UMIs, and fraction of UMIs that originate from mitochondrial genes.

## Filter by Total UMIs
Cells that have too few UMIs per cells as well as those that have too many UMIs per cell are removed from the analysis
```{r total_umi, fig.height=4, fig.width=4, fig.cap="Figure", fig.align="center"}
# determine the sum of UMIs captured for each cell
pData(cells)$total.mRNAs <- Matrix::colSums(exprs(cells))

# determine the average number of UMIs captured per cell
pData(cells)$sample.mean.mRNAs <- mean(pData(cells)$total.mRNAs)

# set the limits
pData(cells)$lower.bound <- pData(cells)$sample.mean.mRNAs / 2.5
pData(cells)$upper.bound <- pData(cells)$sample.mean.mRNAs * 2.5

# identify cells passing the filter
total_mRNA_passed <- row.names(subset(
  pData(cells),
  total.mRNAs > lower.bound[1] & total.mRNAs < upper.bound[1]
))

# plot the distribution of all cells and the filters used
fig.total.mRNA <- ggplot(data = pData(cells)$totalmRNAs, aes(pData(cells)$total.mRNAs)) +
  geom_histogram(binwidth = 200, fill = "#084594") +
  geom_vline(
    xintercept = pData(cells)$lower.bound[1],
    linetype = "dashed", color = "red", size = 0.75
  ) +
  geom_vline(
    xintercept = pData(cells)$upper.bound[1],
    linetype = "dashed", color = "red", size = 0.75
  ) +
  xlab("Total UMIs") +
  ylab("Number of cells") +
  theme_ms() +
  theme(
  
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  )

fig1.b <- fig.total.mRNA
saveManuscriptPlot(fig1.b, 6, 6)
fig1.b
```
## Filter by mtDNA reads
A cell with >= 15% of the total UMIs originating from mitochondrial genes is removed
```{r total_mito, fig.align="center", fig.cap="Figure", fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
# determine the fraction of UMIs from mito mRNAs
pData(cells)$mito.mRNAs <- Matrix::colSums(exprs(cells[row.names(subset(
  fData(cells),
  rownames(fData(cells)) %in% mito.genes
)), ]))

pData(cells)$frac.mito.mRNAs <- pData(cells)$mito.mRNAs / pData(cells)$total.mRNAs # percentage of total UMIs from mtDNA

# filter
mito_mRNA_passed <- row.names(subset(pData(cells), frac.mito.mRNAs < 0.15)) # the cells with < 15% of UMIs from  mtDNA

fig.mtDNA.frac <- pData(cells) %>%
  group_by(as.factor(raw_umi_batch)) %>%
  dplyr::select(frac.mito.mRNAs) %>%
  ggplot(aes(y = frac.mito.mRNAs, x = as.factor(raw_umi_batch), fill = as.factor(raw_umi_batch))) +
  geom_violin(show.legend = FALSE) +
  geom_point(position = "jitter", alpha = 0.2, size = 0.01, color = "black") +
  geom_hline(yintercept = 0.15, linetype = "dashed", color = "red", size = 1) +
  ylab("Mitochondrial UMIs (%)") +
  xlab(" ") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(labels = c(
    "rep1" = "Rep 1", "rep2" = "Rep 2",
    "rep3" = "Rep 3", "rep4" = "Rep 4"
  )) +
  theme_ms() +
  theme(legend.position = "none", axis.text = element_text(size = rel(1)), axis.title = element_text(face = "bold", size = rel(1))) +
  scale_fill_jcolors("pal3")

fig1.c <- fig.mtDNA.frac
saveManuscriptPlot(fig1.c, 6, 6)
fig1.c
```

## Remove poor quality cells
```{r filter_cells, fig.align="center", fig.cap="Figure", fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
# IDs of cells passing both criteria
qc_passed_cells <- intersect(mito_mRNA_passed, total_mRNA_passed)

# retain cells passsing
cells.filter <- cells[, qc_passed_cells]

# plot for each replicate
total_cells <- table(pData(cells)$batch)
qc_passed_cells <- table(pData(cells.filter)$batch)
filtered_cells <- total_cells - qc_passed_cells
df_qc <- t(rbind(qc_passed_cells, filtered_cells))

dat.m <- melt(df_qc)
qc_df <- dat.m %>%
  dplyr::group_by(Var1) %>%
  dplyr::arrange(Var1, desc(Var2)) %>%
  dplyr::mutate(lab_ypos = cumsum(value) - 0.5 * value)

fig.cells.removed <- ggplot(data = qc_df, aes(x = Var1, y = value)) +
  geom_col(aes(fill = Var2), width = 0.7, position = position_stack(reverse = TRUE)) +
  theme(
    legend.title = element_blank(),
    legend.position = "top",
    panel.background = element_blank(),
    axis.line.y = element_line(),
    axis.text.x = element_text(face = "bold")
  ) +
  scale_y_continuous(limits = c(0, 1500)) +
  labs(x = " ", y = "Number of cells") +
  scale_x_discrete(labels = c(
    "rep1" = "Rep 1", "rep2" = "Rep 2",
    "rep3" = "Rep 3", "rep4" = "Rep 4"
  )) +
  theme_ms() +
  theme(legend.position = "none", axis.text = element_text(size = rel(1)), axis.title = element_text(face = "bold", size = rel(1))) +
  theme(
    legend.title = element_blank(),
    legend.position = (c(.75, .94))
    #legend.text = element_text(size = 12)
  ) +
  scale_fill_manual(values = c("#3CA437", "#D7263D"), labels = c("Passed", "Removed"))

fig1.d <- fig.cells.removed
saveManuscriptPlot(fig1.d, 6, 6)
fig1.d
```
## Filter Genes by expression
We now remove genes that are not expressed in more than 50 cells in our dataset or the maximum expression is less than 2 UMIs in any cell
```{r filter_genes, fig.align="center", fig.cap="Figure", fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
cells.filter <- estimateSizeFactors(cells.filter)

cells.filter <- estimateDispersions(cells.filter)

cells.filter <- detectGenes(cells.filter, min_expr = 0.1)

cells.filter <- cells.filter[row.names(subset(
  fData(cells.filter),
  num_cells_expressed > 100
)), ]

fData(cells.filter)$max.expression <- cells.filter %>%
  exprs() %>%
  (qlcMatrix::rowMax) %>%
  as.double()
max.expr.cutoff <- 4

cells.filter <- cells.filter[row.names(subset(
  fData(cells.filter),
  max.expression > max.expr.cutoff
)), ]

# remove mitochondrial genes from the analysis
cells.filter <- cells.filter[!row.names(subset(fData(cells.filter))) %in% mito.genes, ]

fig.genes.expression <- ggplot(data = pData(cells)$num_genes_expressed, aes(pData(cells.filter)$num_genes_expressed)) +
  geom_histogram(binwidth = 15, fill = "#084594") +
  theme_ms() +
  xlab("Detected genes") +
  ylab("Number of cells") +
  theme(
     legend.position = "none",
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  )

fig1.e <- fig.genes.expression
saveManuscriptPlot(fig1.e, 6, 6)
fig1.e
```
### Figure 1

```{r fig.height=6, fig.width=10}


p1 <- (ggplot() + theme_void() + labs(tag="(a)"))/ (ggplot() + theme_void())
p2 <- (ggplot() + theme_void()) / (ggplot() + theme_void())
p3 <- (fig1.b + labs(tag="(b)")) / (fig1.d + labs(tag="(d)"))  
p4 <- (fig1.c + labs(tag="(c)")) /  (fig1.e + labs(tag="(e)")) 

figure1 <- p1 | p2 | p3 | p4
#figure1 <- figure1 + plot_annotation(tag_levels = 'a',tag_prefix = "(", tag_suffix = ")") 
# + plot_annotation(tag_levels = 'a',tag_prefix = "(", tag_suffix = ")") 
saveManuscriptPlot(figure1, 12, 8)
```


## Gene annotation 
The SBG platform utilises the ENSEMBL GTF file for the PICR genome. To improve the annotation of the genes we utlise the corresponding Entrez IDs for those genes labelled with and "ENSCGRG" IDs to obtain a gene symbol
```{r ensembl_annotation}
# set bioMart
ensembl <- useMart(
  host = "http://sep2019.archive.ensembl.org",
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "cgpicr_gene_ensembl"
)

# make a gene_short_name column for the CDS object
fData(cells.filter)$gene_short_name <- rownames(cells.filter)

# how many of the genes have no gene symbol
# sum(str_count(fData(cells.filter)$gene_short_name, "ENSCGR*"))
unannotated <- fData(cells.filter)$gene_short_name[grep("ENSCGR", fData(cells.filter)$gene_short_name)]

# biomart query
annotation <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "entrezgene_id",
    "external_gene_name",
    "description",
    "gene_biotype"
  ),
  filters = c("ensembl_gene_id"),
  values = unannotated,
  mart = ensembl,
  uniqueRows = TRUE
)

# store the final annotation
annotation$final <- "Unannotated"

# 1) add the ensembl annotated external gene names to final columns
# annotation$final[!annotation$external_gene_name == ""] <- annotation$external_gene_name[!annotation$external_gene_name == ""]

# 2) where we have no external gene name from ensembl but we do have an entrez ID
match_entrez_ids <- annotation %>%
  dplyr::filter(!is.na(entrezgene_id))
# load NCBI annotations
cgr_entrez <- readLines("./genelists/cgr_ncbi_ids.txt")

# match entrezID
split_gene_info <- strsplit(cgr_entrez[grepl(paste(match_entrez_ids[, 2], collapse = "|"), cgr_entrez)], "\t")

# extract the matched symbols
entrez_matched <- cbind(
  sapply(split_gene_info, "[", 2),
  sapply(split_gene_info, "[", 3)
)
colnames(entrez_matched) <- c("entrez_gene_id", "symbol")

# add the symbols to the annotation
annotation$final[match(entrez_matched[, 1], annotation$entrezgene_id)] <- entrez_matched[, 2]

# add the new annotations to the CDS object
fData(cells.filter)$gene_short_name[match(annotation$ensembl_gene_id, fData(cells.filter)$gene_short_name)] <- annotation$final
```
# Cell cycle scoring
For the next section we used the cell cycle scoring function of Seurat to determine which phase of the cell cycles cells were in. Using this function will allow us to correct the effect of cell cycle in subsequent analyses.

## Create a Seurat object
```{r create seurat object, message=FALSE, warning=FALSE}
# extract the raw counts from the cds and make a Seurat oject containing the batch variable
cho.obj <- CreateSeuratObject(
  counts = exprs(cells.filter),
  project = "chok1_dp12", min.cells = 3,
  min.features = 200
)

cho.obj <- AddMetaData(
  object = cho.obj,
  metadata = pData(cells.filter)$batch,
  col.name = "batch"
)
```
## Preprocess Seurat 
Before scoring cells by the cell cycle phase any variation due to batch is removed
```{r normalize data in Seurat, message=FALSE, warning=FALSE, include=FALSE}
cho.obj <- NormalizeData(cho.obj)

cho.obj <- FindVariableFeatures(cho.obj,
  selection.method = "vst"
)

cho.obj <- ScaleData(cho.obj,
  features = rownames(cho.obj),
  vars.to.regress = "batch"
)
```
## Specify cell cycle genes
Here we use custom CHO cell genes involved in cell cycle 
```{r convert cell cycle genes to CHO gene ids, message=FALSE, warning=FALSE}
s.genes <- read.table("./genelists/cho.s.txt", stringsAsFactors = F)

g2m.genes <- read.table("./genelists/cho.g2m.txt", stringsAsFactors = F)
```
## Calculate the cell cycle scores
```{r calculate cell cycle scores with Seurat, fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
# run the seurat cell cycle scoring module
cho.obj <- CellCycleScoring(cho.obj,
  s.features = s.genes$V1,
  g2m.features = g2m.genes$V1,
  set.ident = TRUE
)
```
## Add cell cycle scores to CDS
```{r add Seurat cell cycle scores to cds}
pData(cells.filter)$S.Score <- cho.obj$S.Score
pData(cells.filter)$G2M.Score <- cho.obj$G2M.Score
pData(cells.filter)$cc_phase <- cho.obj$Phase
``` 
# tSNE Analysis

## Identify variable genes
```{r select_variable_genes, fig.align="center", fig.cap="Figure", fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
options(warn = -1) # suppress warnings that otherwise clutter output
#' prioritize variable genes for the analysis
disp_table <- dispersionTable(cells.filter)
unsup_clustering_genes <- subset(disp_table, mean_expression > 0.1)
cells.filter <- setOrderingFilter(cells.filter, unsup_clustering_genes$gene_id)

fig.variable.genes <- plot_ordering_genes(cells.filter) +
  theme_ms() +
  theme(aspect.ratio = 1) +
  xlab("Mean Expression") +
  ylab("Dispersion") +
  theme(
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  )

fig2.a <- fig.variable.genes
saveManuscriptPlot(fig2.a, 6, 6)
fig2.a
```
## Select PCs 
```{r PCA_select_plot, fig.align="center", fig.cap="Figure", fig.height=4, fig.width=4, message=TRUE}
fig.pc.variance <- plot_pc_variance_explained(cells.filter, return_all = FALSE) +
  geom_vline(xintercept = 12, linetype = "dashed", color = "red", size = 0.5) +
  labs(x = "Principal Component", y = "Variance Explained") +
  theme_ms() +
  theme(aspect.ratio = 1, axis.text = element_text(size = rel(1)), axis.title = element_text(face = "bold", size = rel(1)))

fig2.b <- fig.pc.variance
saveManuscriptPlot(fig2.b, 6, 6)
fig2.b
```
## t-Stocastic neighbour embedding
```{r tSNE_density_plot, fig.align="center", fig.cap="Figure", fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
cells.filter <- reduceDimension(cells.filter,
  max_components = 2, num_dim = 12,
  reduction_method = "tSNE",
  verbose = FALSE,
  residualModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
  norm_method = "vstExprs"
)
```
## Identify tSNE clusters
```{r}
cells.filter <- clusterCells(cells.filter, rho_threshold = 2, delta_threshold = 4, skip_rho_sigma = T)

fig.scatter.rho.delta <- plot_rho_delta(cells.filter, rho_threshold = 2, delta_threshold = 4) +
  theme_ms() + theme(aspect.ratio = 1)

fig_s_5 <- fig.scatter.rho.delta
saveManuscriptPlot(fig_s_5, 6, 6)
fig_s_5
```
## Effect of CC correction
```{r}
cells.filter <- reduceDimension(cells.filter,
  max_components = 2, num_dim = 12,
  reduction_method = "tSNE",
  verbose = FALSE,
  residualModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs",
  norm_method = "vstExprs"
)

fig.tsne.cc_uncorrected <- plot_cell_clusters(cells.filter, color = "cc_phase", cell_size = 0.8) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_ms() +
  theme(
    legend.position = "right",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1.5))
  )

fig_s_4a <- fig.tsne.cc_uncorrected
saveManuscriptPlot(fig_s_4a, 6, 6)
fig_s_4a

# correct and repeat with Seurat cc_scores
cells.filter <- reduceDimension(cells.filter,
  max_components = 2, num_dim = 12,
  reduction_method = "tSNE",
  verbose = FALSE,
  residualModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
  norm_method = "vstExprs"
)

# show the cell cycle uncorrected
fig.tsne.cc_corrected <- plot_cell_clusters(cells.filter, color = "cc_phase", cell_size = 0.8) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_ms() +
  theme(
    legend.position = "right",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1.5))
  )

fig_s_4b <- fig.tsne.cc_corrected
saveManuscriptPlot(fig_s_4b, 6, 6)
fig_s_4b
```

## tSNE density
```{r fig.height=5, fig.width=5}
# extract the monocle reduced dimesions
tsne_coordinates <- t(reducedDimA(cells.filter))

df <- data.frame(
  x = tsne_coordinates[, 1],
  y = tsne_coordinates[, 2],
  d = densCols(tsne_coordinates[, 1], tsne_coordinates[, 2],
    nbin = 64, bandwidth = 1.5,
    colramp = colorRampPalette(rev(rainbow(10, end = 4 / 6)))
  )
)

fig.tsne.density <- ggplot(df) +
  geom_point(aes(x, y, col = d), size = 0.2) +
  scale_color_identity() +
  theme_ms() +
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  ) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2")

fig2.c <- fig.tsne.density
saveManuscriptPlot(fig2.c, 6, 6)
fig2.c
```

## tSNE clusters
To identify groups  of cells form the tSNE plot we use the louvain method in
monocle with K=100. Each cluster is assinged a color for reference.
```{r cluster_identification, fig.align="center", fig.cap="Figure", fig.height=5, fig.width=5}
fig.tsne.cluster <- plot_cell_clusters(cells.filter, color = "Cluster", cell_size = 0.5) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_ms() +
  theme(
    legend.position = "none",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  ) +
  scale_color_jcolors("pal8")

fig2.d <- fig.tsne.cluster
saveManuscriptPlot(fig2.d, 6, 6)

fig2.d
```
### Figure 2 
```{r}
figure2 <- (fig2.a | fig2.b) / (fig2.c | fig2.d) + 
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
saveManuscriptPlot(figure2, 9, 9)

```


## Differential expression
Using Monocle's differential gene test the genes that differ between the
clusters of cells are identified. We again regress out confounding factors i.e.
batch and cell cycle phase
```{r monocle_cluster_differential_expression, message=TRUE, warning=FALSE}
# monocle differential expression by cluster
tictoc::tic()
cluster_de <- differentialGeneTest(cells.filter,
  fullModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score + Cluster",
  reducedModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
  cores = 32
)
tictoc::toc()
```

# Anti-IL8 mAb gene expression
## bulk RNASeq
```{r mab_expression, fig.height=4, fig.width=10, fig.fullwidth=TRUE, message=FALSE, warning=FALSE}
mab_genes <- c("heavy_chain", "light_chain")

# expression of the mAb genes in bulkRNASeq
bulk_tpm_replicates <- cbind(
  bulk_tpm_r1[bulk_ids],
  bulk_tpm_r2[bulk_ids],
  bulk_tpm_r3[bulk_ids],
  bulk_tpm_r4[bulk_ids]
)
# take the median of the log(TPM+1)
bulk_tpm_median <- data.frame(median_tpm = apply(bulk_tpm_replicates, 1, median))
bulk_tpm_max <- data.frame(median_tpm = apply(bulk_tpm_replicates, 1, max))
bulk_tpm_min <- data.frame(median_tpm = apply(bulk_tpm_replicates, 1, min))

hc_median_tpm <- bulk_tpm_median["heavy_chain", 1]
lc_median_tpm <- bulk_tpm_median["light_chain", 1]
gapdh_median_tpm <- bulk_tpm_median["ENSCGRG00015003621", 1]
actb_median_tpm <- bulk_tpm_median["ENSCGRG00015013883", 1]

fig.density.mab.bulk <- bulk_tpm_median %>%
  ggplot() +
  geom_density(aes(x = median_tpm), fill = "#2E6657", alpha = 0.2) +
  theme_ms() +
  # theme(aspect.ratio = 1) +
  labs(x = "Log (median bulk TPM +1)") +
  geom_segment(aes(x = hc_median_tpm, y = 0, xend = hc_median_tpm, yend = 0.4), linetype = "dashed", color = "#558aa6", arrow = arrow(length = unit(0.15, "cm"), ends = "first", type = "closed")) +
  annotate("text", x = hc_median_tpm, y = 0.42, label = "Heavy chain", color = "#558aa6") +
  geom_segment(aes(x = lc_median_tpm, y = 0, xend = lc_median_tpm, yend = 0.5), linetype = "dashed", color = "#B1740F", arrow = arrow(length = unit(0.15, "cm"), ends = "first", type = "closed")) +
  annotate("text", x = lc_median_tpm + 0.45, y = 0.52, label = "Light chain", color = "#B1740F") +
  geom_segment(aes(x = gapdh_median_tpm, y = 0, xend = gapdh_median_tpm, yend = 0.6), linetype = "dashed", color = "grey", arrow = arrow(length = unit(0.15, "cm"), ends = "first", type = "closed")) +
  annotate("text", x = gapdh_median_tpm, y = 0.62, label = "Gapdh", color = "grey") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))


s_fig5 <- fig.density.mab.bulk
saveManuscriptPlot(s_fig5, 12, 6)
s_fig5
```
## scRNASeq

## HC & LC violin
```{r}
# create a relative expression violin plot for the heavy and light chain
pData(cells.filter)$rel.exprs.hc <- monocle_relative_expression(cells.filter, "heavy_chain")
pData(cells.filter)$rel.exprs.lc <- monocle_relative_expression(cells.filter, "light_chain")
cds_exprs <- cbind(pData(cells.filter)$rel.exprs.hc, pData(cells.filter)$rel.exprs.lc)
cds_exprs <- reshape2::melt(as.matrix(cds_exprs))
colnames(cds_exprs) <- c("Cell", "Gene", "expression")

# log10 adjust the expression
cds_exprs$adjusted_expression <- log10(cds_exprs$expression)

fig.violin.mab <- ggplot(aes_string(x = "Gene", y = "expression", fill = "Gene"), data = cds_exprs) +
  geom_violin(alpha = 0.2) +
  geom_boxplot(width = 0.2) +
  expand_limits(y = c(0.5, 1)) +
  scale_y_log10() +
  scale_fill_jcolors("pal6") +
  theme(legend.position = "none") +
  theme_ms() +
  scale_x_discrete(labels = c("heavy_chain" = "Heavy Chain", "light_chain" = "Light Chain")) +
  # stat_summary(fun.y="median",geom="line",
  #            aes(group="f_id", col="grey"),
  #           position=position_dodge(width= 0.9)) +
  theme(
    legend.position = "none",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  ) +
  labs(y = "Relative expression", x = "") +
  stat_compare_means(method = "wilcox.test", label.x = 0.75)
fig3.b <- fig.violin.mab
saveManuscriptPlot(fig3.b, 6, 6)
fig3.b
```

## HC tSNE
```{r}
fig.tsne.rel.hc <- plot_cell_clusters(cells.filter, color = "rel.exprs.hc", cell_size = 0.5) +
  color.gradient +
  theme_ms() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(face = "bold", size = rel(1))
  ) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  labs(x = "t-SNE 1", y = "t-SNE 2", color = "Relative\nExpression", title = "Heavy Chain")
fig3.d <- fig.tsne.rel.hc
saveManuscriptPlot(fig3.d, 6, 6)
fig3.d
```

## HC cluster expression
```{r}
pData(cells.filter) %>%
  group_by(Cluster) %>%
  summarise(`Cell number` = n(), `Median relative HC expression` = median(rel.exprs.hc), `Max. expression` = max(rel.exprs.hc)) %>%
  DT::datatable(rownames = T,
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(list(
      className = "dt-head-center dt-center",
      targets = 1:2
    ))
  )
) %>% formatSignif(digits = 2, columns = 3:4)

print(paste("Heavy chain DE q-value for cluster = ",signif(cluster_de["heavy_chain",]$qval,2)),sep="")


fig.box.hc.cluster <- pData(cells.filter) %>%
  group_by(Cluster) %>% 
  ggplot(aes(x=Cluster, y = rel.exprs.hc,fill=Cluster)) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.2) +
  scale_y_log10(limits = c(0.5,1e2)) +
  theme_ms() + 
  theme(legend.position = "none", 
        #aspect.ratio = 1, 
        axis.text = element_text(size = rel(1)), 
        axis.title = element_text(face = "bold", size = rel(1)),
        plot.title = element_text(size = rel(1))
  ) +
  labs(y="Relative expression",x="Cluster", title="Heavy Chain") +
  scale_fill_jcolors("pal8")  +
  annotate("text",x="2",y=75, label=expression("DE q-value = "*1.6*" \u00D7 "*10^{-53})) 

fig3.f <- fig.box.hc.cluster
saveManuscriptPlot(fig3.f, 6, 3)
fig3.f
```

## LC tSNE
```{r}
fig.tsne.rel.lc <- plot_cell_clusters(cells.filter, color = "rel.exprs.lc", cell_size = 0.5) + 
 
  color.gradient + 
  theme_ms() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    aspect.ratio = 1,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(face = "bold", size = rel(1))
  ) +
  scale_y_continuous(limits = c(-35, 35)) +
  scale_x_continuous(limits = c(-35, 35)) +
  labs(x = "t-SNE 1", y = "t-SNE 2") +
  labs(x = "t-SNE 1", y = "t-SNE 2", color = "Relative\nExpression", title = "Light Chain")
fig3.e <- fig.tsne.rel.lc
saveManuscriptPlot(fig3.e, 6, 6)
fig3.e
```
## HC cluster expression
```{r}
pData(cells.filter) %>%
  group_by(Cluster) %>%
  summarise(`Cell number` = n(), `Median relative LC expression` = median(rel.exprs.lc), `Max. expression` = max(rel.exprs.lc)) %>%
  DT::datatable(rownames = T,
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(list(
      className = "dt-head-center dt-center",
      targets = 1:2
    ))
  )
) %>% formatSignif(digits = 2, columns = 3:4)

print(paste("Light chain DE q-values for cluster = ",signif(cluster_de["light_chain",]$qval,2)),sep="")


fig.box.lc.cluster <- pData(cells.filter) %>%
  group_by(Cluster) %>% 
  ggplot(aes(x=Cluster, y = rel.exprs.lc,fill=Cluster)) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.2) +
  scale_y_log10(limits = c(0.5,1e2)) +
  theme_ms() + 
  theme(legend.position = "none", 
        #aspect.ratio = 1, 
        axis.text = element_text(size = rel(1)), 
        axis.title = element_text(face = "bold", size = rel(1)),
        plot.title = element_text(size = rel(1))
  ) +
  labs(y="Relative expression",x="Cluster", title="Light Chain") +
  scale_fill_jcolors("pal8")  +
  annotate("text",x="2",y=75, label=expression("DE q-value = "*1.3*" \u00D7 "*10^{-36})) 

fig3.g <- fig.box.lc.cluster
saveManuscriptPlot(fig3.g, 6, 3)
fig3.g
```

### Figure 3
```{r fig.height=9, fig.width=9}
p1 <-  fig3.b | fig2.d + theme(legend.position = "right")
p2 <- fig3.f / fig3.g
p3 <-  fig3.d +theme(legend.position = "right") | fig3.e +theme(legend.position = "right")
  
 
p4 <- p1 / p2 / p3
figure3 <- p4 + plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
saveManuscriptPlot(figure3,9,12)
```


# Trajectory analysis
## Construct trajectory 
using the genes found to vary beteen the identified clusters we first project
onto a 2D space using the DDRTree algorithm before ording the cells in
pseduotime
```{r generate_monocle_trajectory, warning=FALSE, fig.cap="Figure", fig.align="center"}
de_genes <- cluster_de %>%
  arrange(qval) %>%
  filter(qval < 0.01)

cells.filter <- setOrderingFilter(cells.filter, de_genes$type)

cells.filter <- reduceDimension(cells.filter,
  max_components = 2,
  reduction_method = "DDRTree",
  residualModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
  norm_method = "vstExprs"
)

cells.filter <- orderCells(cells.filter)

# cell states are assigned arbitrary numbers, here we reassign #
# to read left to right in the trajectory plot
pData(cells.filter)$State <- mapvalues(pData(cells.filter)$State,
                                      from = c("6", "7", "5", "4", 
                                               "8", "3", "2", "9", "1"),
                                        to = c("1", "2", "3", "4", 
                                               "5", "6", "7", "8", "9")
                                      )

pData(cells.filter)$State <- factor(pData(cells.filter)$State,
                                     levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9")
                                    )

fig.trajectory.state <- plot_cell_trajectory(cells.filter,
  theta = 0,
  show_branch_points = FALSE,
  color_by = "State",
  cell_size = 0.5) +
  labs(x = "Component 1", y = "Component 2", title="Monocle trajectory") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_ms() +
  scale_x_reverse() + # we reverse the direction of the x-axis so that State 1 is at the left
  scale_color_manual(values = colour.set) +
  scale_alpha(pData(cells.filter)$State) +
  theme(
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  )

# modify the transperancy of the cells on the plot
# fig.trajectory.state <- ggplot_build(fig.trajectory.state)
# fig.trajectory.state$data[[2]]$alpha <- 0.3
# 
# fig.trajectory.state <- ggplot_gtable(fig.trajectory.state)
fig5.a <- fig.trajectory.state
saveManuscriptPlot(fig5.a, 8, 4)
fig5.a

# plot the number of cells in each state
fig.bar.state.cells <- pData(cells.filter) %>%
  dplyr::group_by(State) %>%
  dplyr::summarise(num_cells = n()) %>%
  ggplot(aes(x = State, y = num_cells, fill = State)) +
  geom_text(aes(label = num_cells), position = position_dodge(width = 0.9), vjust = -0.25) +
  geom_bar(alpha = 1, stat = "identity") +
  ylim(c(0,1050)) +
  theme(legend.position = "none") +
  theme_ms() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  ) +
  labs(y = "Number of cells", x = "State") +
  scale_fill_manual(values = colour.set)

fig5.b <- fig.bar.state.cells
saveManuscriptPlot(fig5.b, 6, 6)
fig5.b
```
## Median relative HC & LC in each cell state
```{r}
pData(cells.filter) %>%
  group_by(State) %>%
  summarise(`Cell number` = n(), `Median relative HC expression` = median(rel.exprs.hc), `Max. expression` = max(rel.exprs.hc)) %>%
  DT::datatable(rownames = T,
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(list(
      className = "dt-head-center dt-center",
      targets = 1:2
    ))
  )
) %>% formatSignif(digits = 2, columns = 3:4)

pData(cells.filter) %>%
  group_by(State) %>%
  summarise(`Cell number` = n(), `Median relative LC expression` = median(rel.exprs.lc), `Max. expression` = max(rel.exprs.lc)) %>%
  DT::datatable(rownames = T,
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(list(
      className = "dt-head-center dt-center",
      targets = 1:2
    ))
  )
) %>% formatSignif(digits = 2, columns = 3:4)
```


## Differential expression of LC & HC 
```{r}
# j<-1
# for (i in 2:9) {
# state_diff_expression <- list()
# state_diff_expression[[j]]<- differentialGeneTest(cells.filter[, pData(cells.filter)$State %in% c(1, 2)], 
#                                         fullModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score + State",
#                                         reducedModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
#                                         cores = detectCores())
# j<-j+1
# }
```

## HC state expression
```{r}
fig.box.hc.state <- pData(cells.filter) %>%
  group_by(State) %>%
  ggplot(aes(x = State, y = rel.exprs.hc, fill = State)) +
  geom_violin(alpha = 0.2) +
  geom_boxplot(alpha = 1, width = 0.15) +
  scale_y_log10(limits = c(0.5,1e2)) +
  theme_ms() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))
  ) +
  labs(y = "Relative expression", x = "State", title = "Heavy Chain") +
  stat_compare_means(
    label.y.npc = c(0.95),
    label = "p.signif",
    method = "wilcox.test",
    ref.group = "1",
    hide.ns = TRUE,
    size=7
  ) +
  scale_fill_manual(values = colour.set)

fig5.c <- fig.box.hc.state
saveManuscriptPlot(fig5.c, 6, 6)
fig5.c
```

## HC trajectory
```{r message=FALSE, warning=FALSE}
# overlay the heavy chain expression on the trajectory plot
fig.trajectory.rel.hc <- plot_cell_trajectory(cells.filter,
  theta = 0,
  show_branch_points = FALSE,
  color_by = "rel.exprs.hc",
  cell_size = 0.5
) +
  color.gradient +
  labs(x = "Component 1", y = "Component 2", color = "Relative\nExpression", title = "Heavy Chain") +
  theme_ms() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))
  ) +
  scale_x_reverse()

fig5.e <- fig.trajectory.rel.hc
saveManuscriptPlot(fig5.e, 12, 6)
fig5.e
```

## LC state expression
```{r}
# relative expression of the light chain in each cell state
fig.box.lc.state <- pData(cells.filter) %>%
  group_by(State) %>%
  ggplot(aes(x = State, y = rel.exprs.lc, fill = State)) +
  geom_violin(alpha = 0.2) +
  geom_boxplot(alpha = 1, width = 0.15) +
  scale_y_log10(limits = c(0.5,1e2)) +
  theme(legend.position = "none") +
  theme_ms() +
  theme(
    legend.position = "none",
  
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))
  ) +
  labs(y = "Relative expression", x = "State", title = "Light Chain") +
  stat_compare_means(
    label.y.npc = c(0.95),
    label = "p.signif",
    method = "wilcox.test",
    ref.group = "1",
    hide.ns = TRUE, 
    size=7,
    color="black"
  ) +
  scale_fill_manual(values = colour.set)

fig5.d <- fig.box.lc.state
saveManuscriptPlot(fig5.d, 6, 6)
fig5.d
```

## LC trajectory
```{r}
# overlay the light chain expression on the trajectory plot
fig.trajectory.rel.lc <- plot_cell_trajectory(cells.filter,
  theta = 0,
  show_branch_points = FALSE,
  color_by = "rel.exprs.lc",
  cell_size = 0.5
) +
  color.gradient +
  labs(x = "Component 1", y = "Component 2", color = "Relative\nExpression", title = "Light Chain") +
  theme_ms() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))
  ) +
  scale_x_reverse()

fig5.f <- fig.trajectory.rel.lc
saveManuscriptPlot(fig5.f, 12, 6)
fig5.f
```

### Figure 5
Create and save figure 5 in the manuscript.
```{r fig.height=12, fig.width=10}
p1 <- (fig5.a + labs(tag="(a)")) / (fig5.e + labs(tag="(c)")) / (fig5.f + labs(tag="(e)"))   
p2 <-  (fig5.b + labs(tag="(b)")) / (fig5.c + labs(tag="(d)")) / (fig5.d + labs(tag="(f)"))

figure5 <- p1 | p2 + plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
saveManuscriptPlot(figure5,16,11)
```


# Expression of CHO genes along trajectory

## Set the root of the Monocle trajectory
Cell state 1 is set as the root or beginning of the trajectory
```{r traj_plot_pseudotime}
# set the root state
cells.filter <- orderCells(cells.filter, root_state = 1)

fig.trajectory.pseudotime <- plot_cell_trajectory(cells.filter,
  color_by = "Pseudotime",
  theta = 0,
  show_branch_points = FALSE,
  cell_size = 0.5
) +
  labs(x = "Component 1", y = "Component 2", title="Monocle pseudotime") +
  scale_color_viridis_c(option = "D") +
  theme_ms() +
  theme(
    legend.position = "right",
    legend.title.align = 0.5,
    axis.text = element_text(size = rel(1)),
    axis.title = element_text(face = "bold", size = rel(1))
  ) +
  scale_x_reverse()

fig6.a <- fig.trajectory.pseudotime
saveManuscriptPlot(fig6.a, 7, 3)

fig6.a
```
## Pseudotime differential expression
Identification of genes that correlated with Pseudotime. Those genes found to have a qval < 0.01 using the Monocle `differentialGeneTest` function were considered significant
```{r monocle_pseudotime_de, message=TRUE, warning=FALSE, fig.cap="Figure", fig.align="center"}
pseudotime_de <- differentialGeneTest(cells.filter,
  fullModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score + sm.ns(Pseudotime, df = 3)",
  reducedModelFormulaStr = "~batch + num_genes_expressed + total.mRNAs + S.Score + G2M.Score",
  cores = 32
)

# select significant genes
pseudotime_de <- pseudotime_de[pseudotime_de$qval < 0.01, ]

# save the results for Table S3
table.s3 <- pseudotime_de %>%
  arrange(qval) %>%
  dplyr::select(type, gene_short_name, num_cells_expressed, max.expression, pval, qval) %>%
  dplyr::rename(`SBG Annotation` = type, `Gene Symbol` = gene_short_name, `p-value` = pval, `adj p-value` = qval)

tabsdir <- "./tables/"
if (!dir.exists(tabsdir)) {
  dir.create(tabsdir)
}

fn <- paste("tables/Table S3.xlsx", sep = "")
suppressMessages(if (file.exists(fn)) {
  file.remove(fn)
})

write_xlsx(list(pseudotime_de = table.s3),
  path = fn,
  format_headers = TRUE
)
```

## Cluster analysis
```{r trajectory_cluster}
cluster_number <- 5

fig.heatmap.pseudotime <- plot_pseudotime_heatmap(cells.filter[rownames(pseudotime_de), ],
  cores = 32,
  use_gene_short_name = F,
  num_clusters = cluster_number,
  show_rownames = F,
  return_heatmap = T,
  norm_method = "vstExprs"
)

fig6.b <- fig.heatmap.pseudotime
saveManuscriptPlot(fig6.b, 6, 6)
fig6.b
```

## Enrichment analysis  
Use WebGestaltR to determine if the identified clusters of genes are enriched when compared to GO. Those biological processes with a Benjamini Hochberg adjusted p-value < 0.05 were considered significantly enriched.
```{r enrichment_analysis, message=FALSE, warning=FALSE, include=FALSE, paged.print=TRUE}
# extract the genes in each cluster of the Pseudotime heatmap
pseudotime_hm_clusters <- as.data.frame(cutree(fig.heatmap.pseudotime$tree_row, k = cluster_number)) # first cut the dendrogram

pseudotime_hm_clusters$Gene <- rownames(pseudotime_hm_clusters) # get the geneids

pseudotime_hm_clusters$short_name <- fData(cells.filter)$gene_short_name[rownames(fData(cells.filter))
                                      %in% pseudotime_hm_clusters$Gene] # get the corresponding short names

colnames(pseudotime_hm_clusters) <- c("Cluster", "Gene", "short_name")

enrichdir <- "./enrichment_analysis/"
suppressMessages(if (file.exists(enrichdir)) {
  unlink(enrichdir)
})
if (!dir.exists(enrichdir)) {
  dir.create(enrichdir)
}

# write each gene list to file for enrichment analysis
for (i in 1:cluster_number) {
  enrichment_genes <- pseudotime_hm_clusters %>%
    filter(Cluster == i) %>%
    dplyr::select(short_name)
  write(enrichment_genes$short_name, file = paste(enrichdir, "cluster_", i, "_genes.txt", sep = ""))
}

# gene set enrichment analysis
enrich_result <- WebGestaltRBatch(
  enrichMethod = "ORA",
  organism = "mmusculus",
  enrichDatabase = c("geneontology_Biological_Process_noRedundant"),
  enrichDatabaseType = "genesymbol",
  interestGeneFolder = enrichdir,
  interestGeneType = "genesymbol",
  referenceSet = "genome",
  minNum = 10,
  maxNum = 500,
  sigMethod = "fdr",
  fdrMethod = "BH",
  fdrThr = 0.05,
  topThr = 10,
  reportNum = 20,
  perNum = 1000,
  projectName = "Pseudotime",
  isOutput = TRUE,
  outputDirectory = enrichdir,
  dagColor = "continuous",
  setCoverNum = 10,
  networkConstructionMethod = NULL,
  neighborNum = 10,
  highlightType = "Seeds",
  highlightSeedNum = 10,
  nThreads = 32
)

# create the supplementary table S4 for the clusters with enrichment
fn <- paste("tables/Table S4.xlsx", sep = "")
suppressMessages(if (file.exists(fn)) {
  file.remove(fn)
})

write_xlsx(list(
  Cluster1 = enrich_result[[1]]$enrichResult[, c(1, 2, 4, 5, 7, 8, 9, 11)],
  Cluster2 = enrich_result[[2]]$enrichResult[, c(1, 2, 4, 5, 7, 8, 9, 11)]
),
path = fn,
format_headers = TRUE
)
```
## Enriched GO BPs {.tabset .tabset-fade .tabset-pills}
### Cluster 1
```{r cluster1_enrichment}
datatable(enrich_result[[1]]$enrichResult[, c(1, 2, 4, 5, 7, 8, 9)],
  rownames = T, filter = "top",
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(
      list(
        className = "dt-head-center dt-center",
        targets = 3:6
      )
    )
  )
) %>% formatSignif(digits = 2, columns = 5:7)
```
### Cluster 2
```{r cluster2_enrichment}
datatable(enrich_result[[2]]$enrichResult[, c(1, 2, 4, 5, 8, 9)],
  rownames = T,
  filter = "top",
  options = list(
    pagelength = T,
    scrollX = T,
    columnDefs = list(list(
      className = "dt-head-center dt-center",
      targets = 3:6
    ))
  )
) %>% formatSignif(digits = 2, columns = 5:6)
```

## CHO cell gene expression {.tabset .tabset-fade .tabset-pills}
Overlay the expression of genes found to significantly correlate with the progression of cells along the trajectory. 

### Psap
Plot the expression of the **Prosaposin** gene 
```{r}
# relative expression
pData(cells.filter)$rel.exprs.psap <- monocle_relative_expression(cells.filter, "Psap")

fig.traj.rel.psap <- plot_cell_trajectory(cells.filter, 
                                              theta = 0, 
                                              show_branch_points = FALSE, 
                                              color_by = "rel.exprs.psap", 
                                              cell_size = 0.5) +
                                              color.gradient + 
                                              labs(x = "Component 1", 
                                                   y = "Component 2",
                                                   title = "Psap",
                                                   color="Relative\nexpression") +
                                              theme_ms() +
  theme(legend.position = "right",
        axis.text = element_text(size = rel(1)), 
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))) +
  scale_x_reverse()
                                              

fig6.c <- fig.traj.rel.psap
saveManuscriptPlot(fig6.c, 7, 3)
fig6.c
```
### Hmox1
Plot the expression of the **Heme oxygenase 1** gene 
```{r}
# relative expression
pData(cells.filter)$rel.exprs.hmox1 <- monocle_relative_expression(cells.filter, "Hmox1")

fig.traj.rel.hmox1 <- plot_cell_trajectory(cells.filter, 
                                              theta = 0, 
                                              show_branch_points = FALSE, 
                                              color_by = "rel.exprs.hmox1", 
                                              cell_size = 0.5) +
                                              color.gradient + 
                                              labs(x = "Component 1", 
                                                   y = "Component 2",
                                                   title = "Hmox1",
                                                   color="Relative\nexpression") +
                                              theme_ms() +
  theme(legend.position = "right",
        axis.text = element_text(size = rel(1)), 
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))) +
  scale_x_reverse()
                                            

fig6.e <- fig.traj.rel.hmox1
saveManuscriptPlot(fig6.e, 7, 3)
fig6.e
```
### Fth1
Plot the expression of the **Ferritin 1 heavy chain 1** gene 
```{r}
# relative expression
pData(cells.filter)$rel.exprs.fth1 <- monocle_relative_expression(cells.filter, "Fth1")

fig.traj.rel.fth1 <- plot_cell_trajectory(cells.filter, 
                                              theta = 0, 
                                              show_branch_points = FALSE, 
                                              color_by = "rel.exprs.fth1", 
                                              cell_size = 0.5) +
                                              color.gradient + 
                                              labs(x = "Component 1", 
                                                   y = "Component 2",
                                                   title = "Fth1",
                                                   color="Relative\nexpression") +
                                              theme_ms() +
  theme(legend.position = "right",
        axis.text = element_text(size = rel(1)), 
    axis.title = element_text(face = "bold", size = rel(1)),
    plot.title = element_text(size = rel(1))) +
  scale_x_reverse()
                                              

fig6.f <- fig.traj.rel.fth1
saveManuscriptPlot(fig6.f, 7, 3)
fig6.f
```

#### Figure 6
Create and save figure 6 in the manuscript.
```{r}
p1 <- (fig6.a + labs(tag="(a)")) / (ggplot() + theme_void() + labs(tag="(b)")) / (ggplot() + theme_void()) / (ggplot() + theme_void()) / (ggplot() + theme_void())
p2 <- (fig6.f + labs(tag="(c)")) / (fig6.e + labs(tag="(d)")) / (fig6.c + labs(tag="(e)"))

figure6 <- p1 |  p2 

saveManuscriptPlot(figure6,12,10)
```


# Session info
```{r}
devtools::session_info()
```
